\documentclass[12pt, letterpaper]{article}

\usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{enumerate}
\usepackage{newunicodechar}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{multicol}

\DeclareUnicodeCharacter{2212}{-}

\theoremstyle{remark}
\newtheorem{problem}{Problem}

\theoremstyle{remark}
\newtheorem*{solution}{\textbf{Solution}}

\newenvironment{amatrix}[1]{%
	\left[\begin{array}{@{}*{#1}{c}|c@{}}
			}{                 %
		\end{array}\right]
}

\title{Subcubic Parallelized Matrix Inversion \\
	\large Through Strassen's Algorithm}
\author{Andres Valdes, Alejandro Perez, Dewey Mowris, Andy Herrera}
\date{April, 2018}

\begin{document}
    \begin{titlepage}
        \maketitle
        \thispagestyle{empty}

    \end{titlepage}

    \begin{abstract}
        
        Matrix inversion is one of the most computationally complex 
        matrix operations. This is unfortunate, as matrix inversion 
        has many practical applications, including encryption, where the 
        contents of a matrix are obfuscated by multiplying by a square,
        invertible cypher matrix and can only be retrieved my multiplying
        it by the inverse of that cypher.

    \end{abstract}

    \pagebreak

    \section{Introduction}

    \noindent 
    A matrix is a rectangular array of numbers or expressions, for example: 
    \[\begin{bmatrix*}
        a & b & c & d \\
        e & f & g & h
    \end{bmatrix*}\]

    This is what is referred to as a \(2\times{4}\) matrix, as it 
    has two rows and four columns. In general, we speak about matrices as 
    \(m\times{n}\), where \(m\) is the number of rows and \(n\) is the number 
    of columns. The types of matrices we will deal with are called "square" 
    matrices, this is the case that we have an \(m\times{n}\) matrix such that 
    \(m = n\), we refer to these also as \(n\times{n}\) matrices. Only square, 
    non-singular\footnote{A matrix whose determinant is nonzero.} 
    matrices are invertible.

    The inverse of a matrix \(A\), is a matrix \(A^{-1}\) that when multiplied by \(A\), 
    yields the identity matrix, \(I\). \(I_n\) is the square, \(n\times{n}\) 
    matrix with the top-leftmost to the bottom-rightmost diagonal filled in 
    with ones, and where every other element is a zero.
    \[I_n = \begin{bmatrix*}
        1 & 0 & 0 & \dots & 0 \\
        0 & 1 & 0 & \dots & 0 \\
        0 & 0 & 1 & \dots & 0 \\
        \vdots & \vdots & \vdots & \ddots & 0 \\
        0 & 0 & 0 & \dots & 1
    \end{bmatrix*}_{n\times n}\]

    So \(AA^{-1} = I\), this is the linear algebra equivalent of \(a\frac{1}{a} = 1\).

    \bigskip

    Traditionally, the inversion of a matrix involves the use of the formula
    \[A^{-1} = \frac{1}{det(A)}adj(A),\ adj(A) = (C_{A})^T\]

    Where \(C_A\) is the coefficient matrix of \(A\), that is, a matrix of 
    the dimension of \(A\) with determinants filled in for each of the elements.

    This is a particularly heavy computation, as the common algorithm for determinant
    computation is of the computation complexity \(O(n!)\), with the best algorithm 
    being \(O(n^3)\), forcing us to do \(O(n^5)\) computations for the coefficient 
    matrix alone, as there are \(n^2\) elements in \(A\).

    \bigskip

    Thankfully, we do have a way of reducing the problem for the calculation of 
    a matrix inverse via an analytic inversion formula known as blockwise inversion\cite{zhang}. 
    In blockwise inversion, we split our matrix \(M\) into matrix blocks, such that
    \[M^{-1} = \begin{bmatrix*}
        A & B \\
        C & D
    \end{bmatrix*}^{-1} = 
    \begin{bmatrix*}
        A^{-1} + A^{-1}B(D - CA^{-1}B)^{-1}CA^{-1} & -A^{-1}B(D - CA^{-1}B)^{-1} \\
        -(D - CA^{-1}B)^{-1}CA^{-1} & (D - CA^{-1}B)^{-1}
    \end{bmatrix*}\]
    
    The advantage to this divide-and-conquer algorithm is that it can be shown 
    that the computational complexity of the entire inversion is the same as the
    computational complexity of the multiplication algorithm used\cite{algo}. This is where 
    Strassen's algorithm, with a complexity of \(O(n^{log_27})\), comes into play.

    \bigskip

    Strassen describes an algorithm to multiply two matrices by splitting
    each into four matrix blocks of equal dimensions and performing seven multiplications
    between those blocks as opposed to eight\cite{strassen}.
    \[A = \begin{bmatrix*}
        A_{11} & A_{12} \\
        A_{21} & A_{22}
    \end{bmatrix*},\ B =
    \begin{bmatrix*}
        B_{11} & B_{12} \\
        B_{21} & B_{22}
    \end{bmatrix*},\ AB = C =
    \begin{bmatrix*}
        C_{11} & C_{12} \\
        C_{21} & C_{22}
    \end{bmatrix*}\]

    We define seven new matrices
    \begin{align*}
        M_1 &\coloneqq (A_{11} + A_{22})(B_{11} + B_{22}) \\
        M_2 &\coloneqq (A_{21} + A_{22})B_{11} \\
        M_3 &\coloneqq A_{11}(B_{12} - B_{22}) \\
        M_4 &\coloneqq A_{22}(B_{21} - B_{11}) \\
        M_5 &\coloneqq (A_{11} + A_{12})B_{22} \\
        M_6 &\coloneqq (A_{21} - A_{11})(B_{11} + B_{12}) \\
        M_7 &\coloneqq (A_{12} - A_{22})(B_{21} + B_{22})
    \end{align*}

    Note that for each matrix, we multiply only once. We can 
    now compose these matrices into \(C\), or \(AB\).
    \begin{align*}
        C_{11} &= M_1 + M_4 - M_5 + M_7 \\
        C_{12} &= M_3 + M_5 \\
        C_{21} &= M_2 + M_4 \\
        C_{22} &= M_1 - M_2 + M_3 + M_6
    \end{align*}
    And thus, we've achieved matrix multiplication in subcubic time,
    and by extension, we can achieve matrix inversion in subcubic time.

    \section{Previous Works}

    There exist many algorithms for inverting matrices in 
    cubic time via Gaussian Elimination in parallel, similar to the one implemented in many of the popular linear 
    algebra packages like BLAS and LAPACK, and also the parallelized implementations of it.

    There are also parallelized implementations of Strassen's algorithms in BLAS and elsewhere, but their implementation 
    does not strictly tie in with the matrix inversion problem addressed in this article. The topic is particularly sparse 
    when looking towards parallelized inversion algorithms in Go, where parallelized Strassen algorithms are 
    scarse to begin with.
    
    This article aims to implement an algorithm that, while not new, hasn't been parallelized for 
    the application of inverting matrices, especially not in Go.

    \pagebreak

    \section{Process}

    Now, in order to implement the algorithm described above, we'll be using Go, Google's
    language written particularly for concurrency using a wonderful structure called a "goroutine."

    Essentially, our algorithm can be broken up into two algorithms, both of which have aspects that can be 
    parallelized. First and foremost, we need to split our matrix sub-blocks. We then perform 
    multiplications across those sub-blocks using Strassen's algorithm in order to get the seven intermediate matrices, 
    and we do so in parallel. Of course, the intermediate scalar multiplications, additions between matrices, and copying of 
    matrices are parallelized trivially in and of themselves.

    The code on the following page implements the mathematical algorithm described below:
    \begin{align*}
        M_1 &\coloneqq (A_{11} + A_{22})(B_{11} + B_{22}) \\
        M_2 &\coloneqq (A_{21} + A_{22})B_{11} \\
        M_3 &\coloneqq A_{11}(B_{12} - B_{22}) \\
        M_4 &\coloneqq A_{22}(B_{21} - B_{11}) \\
        M_5 &\coloneqq (A_{11} + A_{12})B_{22} \\
        M_6 &\coloneqq (A_{21} - A_{11})(B_{11} + B_{12}) \\
        M_7 &\coloneqq (A_{12} - A_{22})(B_{21} + B_{22})
    \end{align*}
    \pagebreak
    \begin{verbatim}
        wg.Add(7)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.copy(m1, Mult(Add(a11, a22), Add(b11, b22)))
        }(&wg)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.copy(m2, Mult(Add(a21, a22), b11))
        }(&wg)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.copy(m3, Mult(a11, Add(b12, Scale(-1, b22))))
        }(&wg)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.copy(m4, Mult(a22, Add(b21, Scale(-1, b11))))
        }(&wg)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.copy(m5, Mult(Add(a11, a12), b22))
        }(&wg)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.copy(m6, Mult(Add(a21, Scale(-1, a11)), Add(b11, b12)))
        }(&wg)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.copy(m7, Mult(Add(a12, Scale(-1, a22)), Add(b21, b22)))
        }(&wg)
        wg.Wait()
    \end{verbatim}

    \textit{You may notice the use of "WaitGroup", these are essentially pools of 
    goroutines that have yet to finish, the main thread will be halted until 
    these goroutines finish.}

    \pagebreak

    After we've acquired these matrices, we can find compose our result matrix, \(C\), 
    in the following manner: 
    \begin{align*}
        C_{11} &= M_1 + M_4 - M_5 + M_7 \\
        C_{12} &= M_3 + M_5 \\
        C_{21} &= M_2 + M_4 \\
        C_{22} &= M_1 - M_2 + M_3 + M_6
    \end{align*}
    Via the following Go implementation:
    \begin{verbatim}
        wg.Add(4)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.Copy(c11, Add(Add(Add(m1, m4), Scale(-1, m5)), m7))
        }(&wg)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.Copy(c12, Add(m3, m5))
        }(&wg)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.Copy(c21, Add(m2, m4))
        }(&wg)
        go func(wg *sync.WaitGroup) {
            defer wg.Done()
            utils.Copy(c22, Add(Add(Add(m1, Scale(-1, m2)), m3), m6))
        }(&wg)
        wg.Wait()
    \end{verbatim}

    Strassen's multiplication, in our implementation, recurses down to a base case of a trivial \(1\times{1}\) 
    multiplication.

    This is done for every step in our blockwise inversion involving multiplication,
    the implementation of which is far too lengthy and complex to include in this article 
    and would be better understood within the context of the entire package, found on the GitHub repository.\footnote{Repository: https://github.com/SubcubicInversion/implementation}

    Once the algorithm has recursed down to the base case, the \(3\times{3}\) case, the function calls close, 
    fully inverting the matrix.

    \section{Results}

    The following results compare out runtime against that of BLAS, the low-level package
    that handles NumPy's matrix inversion.

    \begin{center}

    
        \begin{tabular}{|c|l|l|}
    
            \hline
            Size & Strassen & BLAS  \\
            \hline\hline
            \(2\times2\) & 44.55ns & 367.89ns \\
            \hline
            \(4\times4\) & 648.49ns & 503.06ns \\
            \hline
            \(8\times8\) & 2.41ms & 1.05ms   \\
            \hline
            \(512\times512\) & 2min, 2s & 12.49ms \\
            \hline
        \end{tabular}
    
    \end{center}

    Clearly, BLAS' matrix inversion implementation, despite its \(O(n^3)\) time complexity, is far faster for matrices of reasonable sizes.
    The reason for this is that the assembly-level implementation of BLAS uses cache tricks to significantly reduce the time required to 
    compute the inverse.

    This advantage of caching over lesser complexity does dwindle as \(n\) approaches infinity, however; but the matrices of that magnitude are far 
    too large for our tests to terminate within a reasonable amount of time.

    Another reason that our algorithm may take longer to execute may be the fact that Go is a garbage-collected language, and through recursive calls, many 
    intermediary structures are created and then simply unreferenced, forcing Go to garbage-collect a large host of memory.

    \pagebreak

    \section{Conclusions}

    While for our tests, the algorithm developed to invert a matrix fell short of the algorithms used commonly 
    as implemented by BLAS and LAPACK, it was the first parallelized algorithm to invert a matrix via Strassen's multiplication 
    implemented in Go. It is interesting to see how, despite Go's low-level nature and the parallelization of an algorithm of lesser 
    complexity, it is still beaten out by clever caching tricks to avoid fully computing the inverse element by element.

    This does not mean that we cannot improve on the algorithm, as it be helpful to continue to improve on the space-complexity of the problem and perhaps write the algorithm on a lower-level 
    language, one that is not garbage-collected, like C or Assembly, as BLAS is written.

    Perhaps in the future, we can run this algorithm on a computer fast enough to terminate within a passable amount of time for 
    very large \(n\), so that we may see the aymptotic relationships between both algorithms.
    
    \begin{thebibliography}{9}
        \bibitem{algo} 
        T. H. Cormen, C. E. Leiserson, R. L. Rivest, C. Stein, 
        \textit{Introduction to Algorithms}, 3rd ed., MIT Press, Cambridge, MA, 2009, ยง28.2.
         
        \bibitem{strassen} 
        Strassen, Volker, 
        \textit{Gaussian Elimination is not Optimal}, Numer. Math. 13, p. 354-356, 1969
         
        \bibitem{zhang} 
        Zhang, F., & Zhang, F. (2005). 
        \textit{The Schur Complement and Its Applications: Numerical Methods and Algorithms}. Boston, MA: Springer US.
    \end{thebibliography}
    


\end{document}
